---
title: "TimingSDS8"
output: html_document
date: '2022-06-30'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source("load_and_annotate_tree.R")
source("project_SDS.R")
source("phylofit_with_offset.R")
library("sigfit")
```

### Overdispersion of mutation counts in normal blood

We estimate the expected overdispersion, here measured as the ratio of observed variance to Poisson variance, as a function of age using burdens from normal blood single cell derived colonies reported in Mitchell et al.
```{r}
dat=read.table("~/Downloads/normal_haematopoiesis-main/3_mutation_telomere_analysis/input/Summary_cut.csv",sep=",",header = TRUE)
res=dat %>% group_by(donor_id,age) %>% summarise(nsub=mean(sub_adj),vsub=var(sub_adj),N=n())
res %>% ggplot(aes(x=age,y=log(vsub/nsub)))+geom_point()+ylab("log(variance/burden)")+geom_smooth(method="lm")
lm=lm(data=res %>% mutate(od=vsub/nsub),log(od)~age)
par(mfrow=c(2,2))
plot(lm)
lod.pred=predict(lm,data.frame(age=25),interval = "confidence")
print(exp(lod.pred))
od.pred=exp(lod.pred[,"fit"]) ## convert back to OD as we did log-linear regression
cat("predicted variance/burden=od=",od.pred)
maxod=exp(lod.pred[,"upr"])
cat("predicted max variance/burden=od=",maxod)
ODF=od.pred
ODF_MAX=maxod
```


##  Retrieve tree data for individuals

```{r}
pdd_cache="../cache/PDD.RDS"
PDD=readRDS(pdd_cache)
```
## Retrieve SNVs called for representative SDS patients of the similar age to  SDS8  (SDS6,7 and 9) ("NormalSBS" signature)

```{r}
normal_snvs=do.call("rbind",
                    lapply(c("SDS6","SDS7","SDS9"),
                           function(x) PDD[[x]]$pdx$dat$details %>% 
                             filter(TYPE=="SNV" & is_localx_excluded==0) %>% 
                             dplyr::select(1:6)))
normal_mut_matrix=with(normal_snvs,get_mut_matrix(Chrom,Pos,Ref,Alt))
```

## Retrieve SNVs that map to after acquistion ("MutantCloneSBS" signature)  and also those on the trunk ("TrunkSBS")
Get all child branches of node 13 (see plot below):

```{r}
tt=plot_tree(PDD$SDS8$pdx$tree_ml,cex.label = 0)
node_labels(tt,b_include_tips = TRUE)
TRUNKNODE=13
nodes=get_all_node_children(TRUNKNODE,PDD$SDS8$pdx$tree_ml)
cat(nodes,"\n")
mutant_clone_snvs=PDD$SDS8$pdx$dat$details %>% filter(TYPE=="SNV" & is_localx_excluded==0 & node %in% nodes )
mt_mut_matrix=with(mutant_clone_snvs,get_mut_matrix(Chrom,Pos,Ref,Alt))
trunk_snvs=PDD$SDS8$pdx$dat$details %>% filter(TYPE=="SNV" & is_localx_excluded==0 & node==TRUNKNODE)
trunk_mut_matrix=with(trunk_snvs,get_mut_matrix(Chrom,Pos,Ref,Alt))
cat("Found",nrow(trunk_snvs),"SNVs in non-masked regions on the trunk\n")
```

## The signatures are clearly quite different

```{r}
mutmat=cbind(normal_mut_matrix,mt_mut_matrix,trunk_mut_matrix)
colnames(mutmat)=c("NormalSBS","MutantCloneSBS","TrunkSBS")
plot_96_profile(mutmat)
csm=cos_sim_matrix(mutmat,mutmat)
print(csm)
```

## Estimate the Proportion of trunk SNVs that carry the "normal" signature
Decompose "TrunkSBS" into a sum of "NormalSBS" and "MutantCloneSBS".  The number trunk variants attributed to "NormalSBS" gives a direct estimate of the mutation burden at transformation and hence an estimate of the timing of transformation:

```{r}
## normalise counts
mutmatn=t(apply(mutmat,1,"/",colSums(mutmat)))
##  Fit sit to signature using SigFit
mcmcfit <- sigfit::fit_signatures(
  counts = t(mutmat[,"TrunkSBS",drop=FALSE]),
  signatures = t(mutmatn[,c("NormalSBS","MutantCloneSBS")]),
  iter = 10000,warmup = 5000,chains = 1,seed = 1234567)
exposures=retrieve_pars(mcmcfit,par="exposures",hpd_prob=0.95)
print(exposures)
## get approx standard error for subsequent use.
se.approx=(exposures$upper_95["NormalSBS"]-exposures$lower_95["NormalSBS"])/(2*1.96)

```

Reconstruction :

```{r sigfit_reconstruction,fig.width=25, fig.height=18.5, out.width='100%', warning=FALSE}
par(mar=c(5,6,6.5,1))
plot_reconstruction(mcmcfit)
```

## Convert the above into an estimate of the mutaton timing

The number trunk variants attributed to "NormalSBS" gives a direct estimate of the mutation burden at transformation and hence an estimate of the timing of transformation.  We account for the intrinsic stochastic number of mutations acquired by a single cell at a given age using a simple ABC model that assumes the number of mutations acquired is negative binomial distributed with a mean that is itself uncertain because the mutation rate differs between individuals (random effect): 

```{r}
SLOPE=15.08998
INTERCEPT=118.93678
N_TRUNK_MUTS=693.5  # Adjusted mutation burden of trunk.
SLOPE_SD=1
PNORMAL_SBS=as.numeric(exposures$mean["NormalSBS"])
PNORMAL_SBS_SE=as.numeric((exposures$upper_95["NormalSBS"]-exposures$lower_95["NormalSBS"])/(2*1.96)) ## approx stand error.

lb.trunk.muts=N_TRUNK_MUTS*as.numeric(exposures$lower_95["NormalSBS"])
ub.trunk.muts=N_TRUNK_MUTS*as.numeric(exposures$upper_95["NormalSBS"])
cat(sprintf("Number of mutations attributed to NormalsSBS %4.1f (%4.1f-%4.1f)",N_TRUNK_MUTS*PNORMAL_SBS,lb.trunk.muts,ub.trunk.muts))
set.seed(1234567)
timing.abc=function( N=1e6,T=25){
  t0=runif(N,0,T)
  lambda=rnorm(N,SLOPE,SLOPE_SD)
  intercept=INTERCEPT
  od=ODF ##  This is age specific estimate based on the normal patients in Mitchell et al.
  nsub=rnbinom(N,mu=lambda*t0+INTERCEPT,size=(lambda*t0+INTERCEPT)/(od-1))
  nsub_sbs=round(N_TRUNK_MUTS*rtrunc(N,"norm",mean=PNORMAL_SBS,sd=PNORMAL_SBS_SE,a=0,b=1))
  data.frame(nsub=nsub,nsub_sbs=nsub_sbs,t0=t0,lambda=lambda)
}
## Unconditional timing ()
test=timing.abc(N=1e7,T=100)
after.rejection.sampling=test %>% filter(nsub_sbs==nsub)
hist(after.rejection.sampling$t0,xlab="Age(Years)",main="Estimated age at transformation event (Unconditional)",breaks=seq(0,50,1))
cat("Number of samples after rejection sampling",dim(after.rejection.sampling)[1],"\n")
print(quantile(after.rejection.sampling$t0,prob=c(0.025,0.5,0.975)))
## Unconditional timing ()
test=timing.abc(N=1e7,T=25)
after.rejection.sampling=test %>% filter(nsub_sbs==nsub)
cat("Number of samples after rejection sampling",dim(after.rejection.sampling)[1],"\n")
hist(after.rejection.sampling$t0,xlab="Age(Years)",main="Estimated age at transformation event (Conditional)",breaks=seq(0,30,0.5))
Q=quantile(after.rejection.sampling$t0,prob=c(0.025,0.5,0.975))
print(Q)
age.lower.bound=Q[1]+0.728268  ## Add in the typical post conception age at birth.
age.median=Q[2]+0.728268  ## Add in the typical post conception age at birth.
```

### Estimation using phylofit

We fit the ultrametric tree to the lowest likely age based on the data.  We introduce dummy samples so that we can localise the transformation event on a branch and fit the model accordingly.  
```{r}
nwstring=sprintf("((z:0,(z2:0,((b:141.0861791,d:113.2765587):50.45165081,((e:171.1492835,(g:164.5258234,(l:146.9439476,h:127.3626109):37.79341449,(i:137.4571178,k:154.9838747):24.59773156):1.256354434):10.00149207,(f:146.7538421,j:134.5303749):20.28401602):43.5048024):%4.3f):0):%4.3f,zeros:0);",N_TRUNK_MUTS*(1-PNORMAL_SBS),N_TRUNK_MUTS*PNORMAL_SBS)
mt=read.tree(text=nwstring)
mt$agedf=data.frame(tip.label=mt$tip.label)
mt$agedf$age=c(age.lower.bound,age.lower.bound+1e-6,rep(25.728268,10),1e-6)
mt$edge.length=round(mt$edge.length)
ft=fit_tree(mt,switch_nodes = c(16),model="poisson_tree")
par(mfcol=c(1,3))
plot_tree(PDD$SDS8$pdx$tree_ml);title("Original Tree")
tt=plot_tree(mt);title("Augmented Tree")
tt=plot_tree(ft$ultratree);title("Fitted time-based tree")
node_labels(tt)
```

We now estimate the selection using phylofit on the clade defined by node 17
```{r}
phylofit_early=fit_clade(ft$ultratree,node=17,nmutcolony = 10,nwtcolony =0,maxt=25.728,maxSYear = 20000,stan.control = list(adapt_delta=0.99))
print(phylofit_early$res)
```

###  Selection estimate based on median age of transformation

```{r}
mt$agedf$age=c(age.median,age.median+1e-6,rep(25.728268,10),1e-6)
ft2=fit_tree(mt,switch_nodes = c(16),model="poisson_tree")
par(mfcol=c(1,2))
tt=plot_tree(ft2$ultratree);title("Fitted time-based tree\n(Median Timing Estimate)",cex.main=0.7)
tt=plot_tree(ft$ultratree);title("Fitted time-based tree\n(Lower Bound Timing Estimate",cex.main=0.7)
```

We now estimate the selection using phylofit on the clade defined by node 17
```{r}
phylofit_median=fit_clade(ft2$ultratree,node=17,nmutcolony = 10,nwtcolony =0,maxt=25.728,maxSYear = 20000,stan.control = list(adapt_delta=0.99))
print(phylofit_median$res)
```

### Conservative estimate of the statistical significance of SDS8 mutation burden
It will be observed that the colonies share a large part of their history. We can conservatively consider the colonies as a single colony and assess the chance of observing such an extreme burden under the null hypothesis that mutations were accrued according to a negative binomial distribution with a mean equal to the expected number of mutations (496) and a variance that is 2.24 times the mean:

```{r}
tips=PDD$SDS8$pdx$meta$clones_short
tree=PDD$SDS8$pdx$tree_ml
burden=nodeHeights(tree)[match(match(tips,tree$tip.label),tree$edge[,2]),2]
hist(burden)
burden=mean(burden)
cat("Mean burden=",burden,"\n")
pn=2*min(pnbinom(burden,mu=496,size=496/(ODF-1),lower.tail = FALSE),
         pnbinom(burden,mu=496,size=496/(ODF-1),lower.tail = TRUE))
cat("Negative Binomial based two sided p-value",pn)
##  Recalculate using the upper end of the predicted over dispersion range 
pn=2*min(pnbinom(burden,mu=496,size=496/(ODF_MAX-1),lower.tail = FALSE),
         pnbinom(burden,mu=496,size=496/(ODF_MAX-1),lower.tail = TRUE))
cat("Negative Binomial based two sided p-value",pn)
```



